dispRity demo

### 2015-09-29 doc/dispRity.demo.Rnw
Updated vignette

\subsection{Getting the data}
You'll need to do this and that

@@

<<Quick go, fig.width=4, fig.height=4, out.width='.4\\linewidth'>>=
## Separate your ordinated data into different series for example
1+1
## Bootstrap this data
1+1
## Calculate a disparity metric from it
1+1
# Summarise and plot the disparity
summary(1)
plot(1)
@

@@

\subsection{Splitting the data}

>>>>

\subsection{Which data?}
You'll need to do this and that
%To develop

<<data, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
## Loading demo and the package data
library(dispRity)

## An ordinated matrix with the tips only
data(BeckLee_mat50)
dim(BeckLee_mat50)
head(BeckLee_mat50[,1:5])

## An ordinated matrix with the tips and the nodes (named nX)
data(BeckLee_mat99)
dim(BeckLee_mat99)
head(BeckLee_mat99[,1:5], 2)
tail(BeckLee_mat99[,1:5], 2)

## A list of first and last occurence data for some fossils
data(BeckLee_ages)
head(BeckLee_ages)

## And a phylogeny:
data(BeckLee_tree)
plot(BeckLee_tree) ; nodelabels(cex=0.8)
@

@@

<<Quick go, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
## Splitting the data
sliced_data <- time.series(BeckLee_mat99, BeckLee_tree, method = "continuous",
    model = "acctran", time = 5, FADLAD = BeckLee_ages)

## Bootstrapping the data
bootstrapped_data <- boot.matrix(sliced_data, 100)

## Calculating disparity
sum_of_ranges <- dispRity(bootstrapped_data, metric = c(sum, range))

## Summarising the results
summary(sum_of_ranges)
plot(sum_of_ranges, type = "continuous")
@

@@

\subsection{The \dispRity objects}
Disparity analysis involve a lot of shuffling around with many matrices (especially when bootstrapping the data) which can be a bit impractical to visualise and quickly jam your \R console.
For example, try to use \texttt{str} to show the structure of the object created above:

<<dispRity class1, eval=FALSE>>=
str(sum_of_ranges)
@

Therefore this package proposes a new class of object called \dispRity (how unexpected!) that allow to use a \texttt{S3 print.dispRity} method.
In other words, when you are creating objects using the \dispRity functions, you can quickly and easily have a look at them by just printing it.
For example:

<<dispRity class2>>=
## Which class is the sum_of_ranges object?
class(sum_of_ranges)

## We can summarise it using the S3 method print.dispRity
sum_of_ranges 
@

Note that each object output from different functions will show summarise slightly different informations.

\subsection{Splitting the data}
One of the first functionality of this package is to facilitate splitting the ordinated matrix in subsamples.
The original ordinated matrix represents the total any-o-space.

\subsubsection{Custom splitting}
\texttt{cust.series} is a fairly straightforward function and allows to split the data according to a factor determined by the user.
For example, here we can split the matrix based on phylogeny: let's classify in two different groups the crown and stem mammals present in the tree.

<<cust.series2>>=
## We want to separate the species around the node 71 (see phylogeny above).
## All the descendant are crown and all the ancestors are stem
crown <- extract.clade(BeckLee_tree, node = 71)$tip.label
stem <- drop.tip(BeckLee_tree, tip = crown)$tip.label

## We then have to feed this information in a data.frame with one column
factors <- as.data.frame(
    matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem))),
    ncol = 1, dimnames = list(c(crown, stem))))

## Then we can use the customised series function
crown_stem <- cust.series(BeckLee_mat50, factors)

## This created a dispRity object containing two series: crown and stem
class(crown_stem)
crown_stem

## This object contains three elements
names(crown_stem)

## With "data" being the list sub-matrices
str(crown_stem$data)

## "taxa" being the list of taxa in the original ordinated matrix
str(crown_stem$taxa)

## "series" containing information on the series type (custom) and names (crown and stem)
crown_stem$series
@

Of course, using phylogeny as a factor is just an example.
Any type of factors can be used in this example (e.g. body mass categories, habitat, diet, etc...).
Which leads us to another type of category used a lot in disparity-through-time analysis: time!

\subsubsection{Time splitting}
\texttt{time.series} is a function that allows to split the matrix in different ways using time as a category.
Two types of time series splitting can be performed with the \dispRity package by using the \texttt{method} option in \texttt{time.series}.
\begin{enumerate}
\item the classical discrete time splitting (or time-binning) using \texttt{method = "discrete"};
\item or a new continuous time splitting (or time-slicing) using \texttt{method = "continuous"}.
\end{enumerate}

### 2015-09-29 doc/dispRity.demo.tex
Updated vignette

\subsection{Getting the data}
You'll need to do this and that

\section{A super quick go through}
This package allows to:
\begin{enumerate}
    \item Separate your ordinated data into different series for example
    \item Bootstrap this data
    \item Calculate a disparity metric from it
    \item Summarise and plot the disparity
\end{enumerate}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## Separate your ordinated data into different series for example}
\hlnum{1}\hlopt{+}\hlnum{1}
\end{alltt}
\begin{verbatim}
## [1] 2
\end{verbatim}
\begin{alltt}
\hlcom{## Bootstrap this data}
\hlnum{1}\hlopt{+}\hlnum{1}
\end{alltt}
\begin{verbatim}
## [1] 2
\end{verbatim}
\begin{alltt}
\hlcom{## Calculate a disparity metric from it}
\hlnum{1}\hlopt{+}\hlnum{1}
\end{alltt}
\begin{verbatim}
## [1] 2
\end{verbatim}
\begin{alltt}
\hlcom{# Summarise and plot the disparity}
\hlkwd{summary}\hlstd{(}\hlnum{1}\hlstd{)}
\end{alltt}
\begin{verbatim}
##    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
##       1       1       1       1       1       1
\end{verbatim}
\begin{alltt}
\hlkwd{plot}\hlstd{(}\hlnum{1}\hlstd{)}
\end{alltt}
\end{kframe}

{\centering \includegraphics[width=.4\linewidth]{figure/minimal-Quick_go-1}

@@

>>>>

\subsection{Which data?}
You'll need to do this and that
%To develop

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## Loading demo and the package data}
\hlkwd{library}\hlstd{(dispRity)}
\end{alltt}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Loading required package: paleotree}}\begin{alltt}
\hlcom{## An ordinated matrix with the tips only}
\hlkwd{data}\hlstd{(BeckLee_mat50)}
\hlkwd{dim}\hlstd{(BeckLee_mat50)}
\end{alltt}
\begin{verbatim}
## [1] 50 48
\end{verbatim}
\begin{alltt}
\hlkwd{head}\hlstd{(BeckLee_mat50[,}\hlnum{1}\hlopt{:}\hlnum{5}\hlstd{])}
\end{alltt}
\begin{verbatim}
##                    [,1]          [,2]        [,3]       [,4]      [,5]
## Cimolestes   -0.5319679  0.1117759259  0.09865194 -0.1933148 0.2035833
## Maelestes    -0.4087147  0.0139690317  0.26268300  0.2297096 0.1310953
## Batodon      -0.6923194  0.3308625215 -0.10175223 -0.1899656 0.1003108
## Bulaklestes  -0.6802291 -0.0134872777  0.11018009 -0.4103588 0.4326298
## Daulestes    -0.7386111  0.0009001369  0.12006449 -0.4978191 0.4741342
## Uchkudukodon -0.5105254 -0.2420633915  0.44170317 -0.1172972 0.3602273
\end{verbatim}
\begin{alltt}
\hlcom{## An ordinated matrix with the tips and the nodes (named nX)}
\hlkwd{data}\hlstd{(BeckLee_mat99)}
\hlkwd{dim}\hlstd{(BeckLee_mat99)}
\end{alltt}
\begin{verbatim}
## [1] 99 97
\end{verbatim}
\begin{alltt}
\hlkwd{head}\hlstd{(BeckLee_mat99[,}\hlnum{1}\hlopt{:}\hlnum{5}\hlstd{],} \hlnum{2}\hlstd{)}
\end{alltt}
\begin{verbatim}
##                  [,1]       [,2]       [,3]       [,4]        [,5]
## Cimolestes -0.6082437 -0.0323683 0.08458885 -0.4338448 -0.30536875
## Maelestes  -0.5730206 -0.2840361 0.01308847 -0.1258848  0.06123611
\end{verbatim}
\begin{alltt}
\hlkwd{tail}\hlstd{(BeckLee_mat99[,}\hlnum{1}\hlopt{:}\hlnum{5}\hlstd{],} \hlnum{2}\hlstd{)}
\end{alltt}
\begin{verbatim}
##            [,1]      [,2]       [,3]       [,4]       [,5]
## n48 -0.05529018 0.4799330 0.04118477 0.04944912 -0.3558830
## n49 -0.13067785 0.4478168 0.11956268 0.13800340 -0.3222785
\end{verbatim}
\begin{alltt}
\hlcom{## A list of first and last occurence data for some fossils}
\hlkwd{data}\hlstd{(BeckLee_ages)}
\hlkwd{head}\hlstd{(BeckLee_ages)}
\end{alltt}
\begin{verbatim}
##             FAD  LAD
## Adapis     37.2 36.8
## Asioryctes 83.6 72.1
## Leptictis  33.9 33.3
## Miacis     49.0 46.7
## Mimotona   61.6 59.2
## Notharctus 50.2 47.0
\end{verbatim}
\begin{alltt}
\hlcom{## And a phylogeny:}
\hlkwd{data}\hlstd{(BeckLee_tree)}
\hlkwd{plot}\hlstd{(BeckLee_tree) ;} \hlkwd{nodelabels}\hlstd{(}\hlkwc{cex}\hlstd{=}\hlnum{0.8}\hlstd{)}
\end{alltt}
\end{kframe}

{\centering \includegraphics[width=.8\linewidth]{figure/minimal-data-1}

@@

\section{A super quick go through}
This package allows to:
\begin{enumerate}
    \item Splitting the data
    \item Bootstrapping the data
    \item Calculating disparity
    \item Summarising the results
\end{enumerate}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## Splitting the data}
\hlstd{sliced_data} \hlkwb{<-} \hlkwd{time.series}\hlstd{(BeckLee_mat99, BeckLee_tree,} \hlkwc{method} \hlstd{=} \hlstr{"continuous"}\hlstd{,}
    \hlkwc{model} \hlstd{=} \hlstr{"acctran"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlnum{5}\hlstd{,} \hlkwc{FADLAD} \hlstd{= BeckLee_ages)}
\end{alltt}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Some tips have FAD/LAD and are assumed to interval single points in time.}}\begin{alltt}
\hlcom{## Bootstrapping the data}
\hlstd{bootstrapped_data} \hlkwb{<-} \hlkwd{boot.matrix}\hlstd{(sliced_data,} \hlnum{100}\hlstd{)}

\hlcom{## Calculating disparity}
\hlstd{sum_of_ranges} \hlkwb{<-} \hlkwd{dispRity}\hlstd{(bootstrapped_data,} \hlkwc{metric} \hlstd{=} \hlkwd{c}\hlstd{(sum, range))}

\hlcom{## Summarising the results}
\hlkwd{summary}\hlstd{(sum_of_ranges)}
\end{alltt}
\begin{verbatim}
##      series  n   mean   2.5%    25%    75%  97.5%
## 1 133.51104  3 -0.348 -0.397 -0.350 -0.349 -0.230
## 2 100.13328 14  0.101 -0.252 -0.046  0.214  0.291
## 3  66.75552 23 -0.059 -0.163 -0.112  0.031  0.115
## 4  33.37776 11  0.008 -0.134 -0.086  0.132  0.217
## 5         0 10  0.009 -0.134 -0.086  0.140  0.187
\end{verbatim}
\begin{alltt}
\hlkwd{plot}\hlstd{(sum_of_ranges,} \hlkwc{type} \hlstd{=} \hlstr{"continuous"}\hlstd{)}
\end{alltt}
\end{kframe}

{\centering \includegraphics[width=.8\linewidth]{figure/minimal-Quick_go-1}

\section{In more details}

\subsection{The \dispRity objects}
Disparity analysis involve a lot of shuffling around with many matrices (especially when bootstrapping the data) which can be a bit impractical to visualise and quickly jam your \R console.
For example, try to use \texttt{str} to show the structure of the object created above:

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{str}\hlstd{(sum_of_ranges)}
\end{alltt}
\end{kframe}
\end{knitrout}

Therefore this package proposes a new class of object called \dispRity (how unexpected!) that allow to use a \texttt{S3 print.dispRity} method.
In other words, when you are creating objects using the \dispRity functions, you can quickly and easily have a look at them by just printing it.
For example:

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## Which class is the sum_of_ranges object?}
\hlkwd{class}\hlstd{(sum_of_ranges)}
\end{alltt}
\begin{verbatim}
## [1] "dispRity"
\end{verbatim}
\begin{alltt}
\hlcom{## We can summarise it using the S3 method print.dispRity}
\hlstd{sum_of_ranges}
\end{alltt}
\begin{verbatim}
## Disparity measurements across 5 series for 99 taxa. 
## Series:
## 133.51104, 100.13328, 66.75552, 33.37776, 0.
## Disparity calculated as: sum range for 97 dimensions.
## Data was split using continuous method.
## Data was bootstrapped 100 times, using the full bootstrap method.
\end{verbatim}
\end{kframe}
\end{knitrout}

Note that each object output from different functions will show summarise slightly different informations.

\subsection{Splitting the data}
One of the first functionality of this package is to facilitate splitting the ordinated matrix in subsamples.
The original ordinated matrix represents the total any-o-space.

\subsubsection{Custom splitting}
\texttt{cust.series} is a fairly straightforward function and allows to split the data according to a factor determined by the user.
For example, here we can split the matrix based on phylogeny: let's classify in two different groups the crown and stem mammals present in the tree.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## We want to separate the species around the node 71 (see phylogeny above).}
\hlcom{## All the descendant are crown and all the ancestors are stem}
\hlstd{crown} \hlkwb{<-} \hlkwd{extract.clade}\hlstd{(BeckLee_tree,} \hlkwc{node} \hlstd{=} \hlnum{71}\hlstd{)}\hlopt{$}\hlstd{tip.label}
\hlstd{stem} \hlkwb{<-} \hlkwd{drop.tip}\hlstd{(BeckLee_tree,} \hlkwc{tip} \hlstd{= crown)}\hlopt{$}\hlstd{tip.label}

\hlcom{## We then have to feed this information in a data.frame with one column}
\hlstd{factors} \hlkwb{<-} \hlkwd{as.data.frame}\hlstd{(}
    \hlkwd{matrix}\hlstd{(}\hlkwc{data} \hlstd{=} \hlkwd{c}\hlstd{(}\hlkwd{rep}\hlstd{(}\hlstr{"crown"}\hlstd{,} \hlkwd{length}\hlstd{(crown)),} \hlkwd{rep}\hlstd{(}\hlstr{"stem"}\hlstd{,} \hlkwd{length}\hlstd{(stem))),}
    \hlkwc{ncol} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{dimnames} \hlstd{=} \hlkwd{list}\hlstd{(}\hlkwd{c}\hlstd{(crown, stem))))}

\hlcom{## Then we can use the customised series function}
\hlstd{crown_stem} \hlkwb{<-} \hlkwd{cust.series}\hlstd{(BeckLee_mat50, factors)}

\hlcom{## This created a dispRity object containing two series: crown and stem}
\hlkwd{class}\hlstd{(crown_stem)}
\end{alltt}
\begin{verbatim}
## [1] "dispRity"
\end{verbatim}
\begin{alltt}
\hlstd{crown_stem}
\end{alltt}
\begin{verbatim}
## 2 custom series for 50 taxa. 
## Series:
## crown, stem.
\end{verbatim}
\begin{alltt}
\hlcom{## This object contains three elements}
\hlkwd{names}\hlstd{(crown_stem)}
\end{alltt}
\begin{verbatim}
## [1] "data"   "taxa"   "series"
\end{verbatim}
\begin{alltt}
\hlcom{## With "data" being the list sub-matrices}
\hlkwd{str}\hlstd{(crown_stem}\hlopt{$}\hlstd{data)}
\end{alltt}
\begin{verbatim}
## List of 2
##  $ crown: num [1:30, 1:48] 0.3079 0.6531 0.5089 -0.1652 -0.0419 ...
##   ..- attr(*, "dimnames")=List of 2
##   .. ..$ : chr [1:30] "Dasypodidae" "Bradypus" "Myrmecophagidae" "Todralestes" ...
##   .. ..$ : NULL
##  $ stem : num [1:20, 1:48] -0.739 -0.68 -0.511 -0.477 -0.473 ...
##   ..- attr(*, "dimnames")=List of 2
##   .. ..$ : chr [1:20] "Daulestes" "Bulaklestes" "Uchkudukodon" "Kennalestes" ...
##   .. ..$ : NULL
\end{verbatim}
\begin{alltt}
\hlcom{## "taxa" being the list of taxa in the original ordinated matrix}
\hlkwd{str}\hlstd{(crown_stem}\hlopt{$}\hlstd{taxa)}
\end{alltt}
\begin{verbatim}
##  chr [1:50] "Cimolestes" "Maelestes" "Batodon" "Bulaklestes" "Daulestes" ...
\end{verbatim}
\begin{alltt}
\hlcom{## "series" containing information on the series type (custom) and names (crown and stem)}
\hlstd{crown_stem}\hlopt{$}\hlstd{series}
\end{alltt}
\begin{verbatim}
## [1] "custom" "crown"  "stem"
\end{verbatim}
\end{kframe}
\end{knitrout}

Of course, using phylogeny as a factor is just an example.
Any type of factors can be used in this example (e.g. body mass categories, habitat, diet, etc...).
Which leads us to another type of category used a lot in disparity-through-time analysis: time!

\subsubsection{Time splitting}
\texttt{time.series} is a function that allows to split the matrix in different ways using time as a category.
Two types of time series splitting can be performed with the \dispRity package by using the \texttt{method} option in \texttt{time.series}.
\begin{enumerate}
\item the classical discrete time splitting (or time-binning) using \texttt{method = "discrete"};
\item or a new continuous time splitting (or time-slicing) using \texttt{method = "continuous"}.
\end{enumerate}

\subsection{Bootstrapping the data}

\subsection{Calculating disparity}

\subsection{Summarising the results}
